MSDS6372Project2

### 2021-03-23 MSDS6372Project2.Rmd
objective 2 and beyond

# Generate ROC curve to find a cutoff to use
library(ROCR)
results.sm1 <-prediction(fit.pred.sm1, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.sm1 = performance(results.sm1, measure = "tpr", x.measure = "fpr")
plot(roc.sm1,colorize = TRUE)
abline(a=0, b= 1)

# Since this dataset is to predict whether income is less than or equal to 50K or greater than 50K
# We want the sensitivity and specificity to be roughly the same - no reasoon to prefer one group or the other
# As we can see from the color legend, the greant to light blue region will give over 80% True positive rate and less than 20% False positive rate
# And that cut off is roughly correspond to 0.275

@@

# Generate ROC curve to find a cutoff to use
results.complex1 <-prediction(fit.pred.complex1, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.complex1 = performance(results.complex1, measure = "tpr", x.measure = "fpr")
plot(roc.complex1,colorize = TRUE)
abline(a=0, b= 1)
#Calculate logistic regression complex using cut of 0.3 on test set
cutoff1<-0.275
class.complex1 <-factor(ifelse(fit.pred.complex1>cutoff1,">50K.","<=50K."),levels=c("<=50K.",">50K."))
confusionMatrix(class.complex1,adult.test$income)
# The complex model has about the same performance by adding in relationship:marital.status interaction
# Accuracy : 0.817  Sensitivity : 0.8192 Specificity : 0.8102

@@

# Generate ROC curve to find a cutoff to use
results.step1 <-prediction(fit.pred.step1, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.step1 = performance(results.step1, measure = "tpr", x.measure = "fpr")
plot(roc.step1,colorize = TRUE)
abline(a=0, b= 1)
#Calculate logistic regression complex using cut of 0.3 on test set
cutoff1<-0.275
class.step1 <-factor(ifelse(fit.pred.step1>cutoff1,">50K.","<=50K."),levels=c("<=50K.",">50K."))
confusionMatrix(class.step1,adult.test$income)
# The step1 model has about the same performance compared to complex model with interaction
# Accuracy : 0.8174 Sensitivity : 0.8194 Specificity : 0.8112

@@

# Generate ROC curve to find a cutoff to use
results.complex2 <-prediction(fit.pred.complex2, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.complex2 = performance(results.complex2, measure = "tpr", x.measure = "fpr")
plot(roc.complex2,colorize = TRUE)
abline(a=0, b= 1)
#Calculate logistic regression complex using cut of 0.3 on test set
cutoff1<-0.275
class.complex2 <-factor(ifelse(fit.pred.complex2>cutoff1,">50K.","<=50K."),levels=c("<=50K.",">50K."))
confusionMatrix(class.complex2,adult.test$income)
# The complex model has about the same performance by adding in relationship:marital.status interaction
# Accuracy : Accuracy : 0.8167  Sensitivity : 0.8183 Specificity : 0.8118

@@

>>>>

# Generate ROC curve to find a cutoff to use
library(ROCR)
results.sm1 <-prediction(fit.pred.sm1, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.sm1 = performance(results.sm1, measure = "tpr", x.measure = "fpr")
plot(roc.sm1,colorize = TRUE)
abline(a=0, b= 1)
auc.sm1 <- performance(results.sm1, measure = "auc")
auc.sm1 <- auc.sm1@y.values
text(x = .40, y = .6,paste("AUC = ", round(auc.sm1[[1]],3), sep = ""))
# AUC = 0.903
# Since this dataset is to predict whether income is less than or equal to 50K or greater than 50K
# We want the sensitivity and specificity to be roughly the same - no reasoon to prefer one group or the other
# As we can see from the color legend, the greant to light blue region will give over 80% True positive rate and less than 20% False positive rate
# And that cut off is roughly correspond to 0.275

@@

# Generate ROC curve to find a cutoff to use
results.complex1 <-prediction(fit.pred.complex1, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.complex1 = performance(results.complex1, measure = "tpr", x.measure = "fpr")
plot(roc.complex1,colorize = TRUE)
abline(a=0, b= 1)
auc.complex1 <- performance(results.complex1, measure = "auc")
auc.complex1 <- auc.complex1@y.values
text(x = .40, y = .6,paste("AUC = ", round(auc.complex1[[1]],3), sep = ""))
# AUC = 0.903
#Calculate logistic regression complex using cut of 0.3 on test set
cutoff1<-0.275
class.complex1 <-factor(ifelse(fit.pred.complex1>cutoff1,">50K.","<=50K."),levels=c("<=50K.",">50K."))
confusionMatrix(class.complex1,adult.test$income)
# The complex model has about the same performance by adding in relationship:marital.status interaction
# Accuracy : 0.817  Sensitivity : 0.8192 Specificity : 0.8102

@@

# Generate ROC curve to find a cutoff to use
results.step1 <-prediction(fit.pred.step1, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.step1 = performance(results.step1, measure = "tpr", x.measure = "fpr")
plot(roc.step1,colorize = TRUE)
abline(a=0, b= 1)
auc.step1 <- performance(results.step1, measure = "auc")
auc.step1 <- auc.step1@y.values
text(x = .40, y = .6,paste("AUC = ", round(auc.step1[[1]],3), sep = ""))
# AUC = 0.904
#Calculate logistic regression complex using cut of 0.3 on test set
cutoff1<-0.275
class.step1 <-factor(ifelse(fit.pred.step1>cutoff1,">50K.","<=50K."),levels=c("<=50K.",">50K."))
confusionMatrix(class.step1,adult.test$income)
# The step1 model has about the same performance compared to complex model with interaction
# Accuracy : 0.8174 Sensitivity : 0.8194 Specificity : 0.8112

@@

# Generate ROC curve to find a cutoff to use
results.complex2 <-prediction(fit.pred.complex2, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.complex2 = performance(results.complex2, measure = "tpr", x.measure = "fpr")
plot(roc.complex2,colorize = TRUE)
abline(a=0, b= 1)
auc.complex2 <- performance(results.complex2, measure = "auc")
auc.complex2 <- auc.complex2@y.values
text(x = .40, y = .6,paste("AUC = ", round(auc.complex2[[1]],3), sep = ""))
# AUC = 0.904
#Calculate logistic regression complex using cut of 0.275 on test set
cutoff1<-0.275
class.complex2 <-factor(ifelse(fit.pred.complex2>cutoff1,">50K.","<=50K."),levels=c("<=50K.",">50K."))
confusionMatrix(class.complex2,adult.test$income)
# The complex model has about the same performance by adding in relationship:marital.status interaction
# Accuracy : Accuracy : 0.8167  Sensitivity : 0.8183 Specificity : 0.8118

@@

#### Objective 2 Part 2 - Create competing model using continuous predictors and use LDA or QDA
LDA model is the winner after comparison:
Even QDA have slightly better AUC value(suspect QDA is overfit) - the cut off of probability is extremely sensitive(highly dependent on test set)
Thus LDA is the winner
However, LDA performance isn't as good compared to Logistic Regression

```{r}
# construct the LDA model
# Fron EDA we know fnlwgt is not helpful in identifying income - will not use it here
# We will use education.num since it is a substitute for education
mylda <- lda(income ~ age + education.num+ capital.gain+capital.loss+hours.per.week, data = adult)
mylda.prd <- predict(mylda, newdata = adult.test)
mylda.posterior <- mylda.prd$posterior[,2]

# Generate ROC curve to find a cutoff to use
results.lda <-prediction(mylda.posterior, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.lda = performance(results.lda, measure = "tpr", x.measure = "fpr")
plot(roc.lda,colorize = TRUE)
abline(a=0, b= 1)
auc.lda <- performance(results.lda, measure = "auc")
auc.lda <- auc.lda@y.values
text(x = .40, y = .6,paste("AUC = ", round(auc.lda[[1]],3), sep = ""))
# As we can see the lda model is slighly less good compared to logistic regression model
# The ROC curve has shifted downwards compared to LR - AUC =0.809(decreased)
# Nevertheless to balance off performance between high True positive and low False positive we can use around 0.25

#Calculate logistic regression complex using cut of 0.25 on test set
cutoff1<-0.23
class.lda <-factor(ifelse(mylda.posterior>cutoff1,">50K.","<=50K."),levels=c("<=50K.",">50K."))
confusionMatrix(class.lda,adult.test$income)
# Accuracy : Accuracy : 0.726  Sensitivity : 0.7263 Specificity : 0.7251

##################################################################################################
# construct the QDA model
# Fron EDA we know fnlwgt is not helpful in identifying income - will not use it here
# We will use education.num since it is a substitute for education
myqda <- qda(income ~ age + education.num+ capital.gain+capital.loss+hours.per.week, data = adult)
myqda.prd <- predict(myqda, newdata = adult.test)
myqda.posterior <- myqda.prd$posterior[,2]

# Generate ROC curve to find a cutoff to use
results.qda <-prediction(myqda.posterior, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.qda = performance(results.qda, measure = "tpr", x.measure = "fpr")
plot(roc.qda,colorize = TRUE)
abline(a=0, b= 1)
auc.qda <- performance(results.qda, measure = "auc")
auc.qda <- auc.qda@y.values
text(x = .40, y = .6,paste("AUC = ", round(auc.qda[[1]],3), sep = ""))
# AUC seem to be slightly better than LDA AUC = 0.823
# To balance off performance between high True positive and low False positive we can use around 0.25

#Calculate logistic regression complex using cut of 0.015 on test set
cutoff2<-0.015
class.qda <-factor(ifelse(myqda.posterior>cutoff2,">50K.","<=50K."),levels=c("<=50K.",">50K."))
confusionMatrix(class.qda,adult.test$income)
# Accuracy : Accuracy : 0.7461  Sensitivity : 0.7563 Specificity : 0.7150
# Due to the extreme cut off we know even QDA has better AUC value, the prediction cut off is extremely sensitive based on the test set
# or future validation set
# LDA model is much more stable even the AUC value is slightly less

#### Objective 2 part 3 - Nonparametric model approach - Random forest
Default RF ntree = 10 and m variables =sqrt(p) seem to work better than ntree=100
```{r}
library(randomForest)
set.seed(1)
# By default use sqrt(p) number variables when building a random forest of classification trees
rf.adult = randomForest(income~.,data=adult)

# Fix variable levels in adult.test in order to use randomForest
levels(adult.test$workclass)<-levels(adult$workclass)
levels(adult.test$education)<-levels(adult$education)
levels(adult.test$marital.status)<-levels(adult$marital.status)
levels(adult.test$occupation)<-levels(adult$occupation)
levels(adult.test$relationship)<-levels(adult$relationship)
levels(adult.test$race)<-levels(adult$race)
levels(adult.test$native.country)<-levels(adult$native.country)

importance(rf.adult)
plot(rf.adult)
# Run randomForest prediction
testdata = adult.test[,-15]
rf.pred.prob = predict(rf.adult,newdata=testdata,type = "prob")
rf.prob = rf.pred.prob[,2]

# Generate ROC curve to find a cutoff to use
results.rf <-prediction(rf.prob, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.rf = performance(results.rf, measure = "tpr", x.measure = "fpr")
plot(roc.rf,colorize = TRUE)
abline(a=0, b= 1)
auc.rf <- performance(results.rf, measure = "auc")
auc.rf <- auc.rf@y.values
text(x = .40, y = .6,paste("AUC = ", round(auc.rf[[1]],3), sep = ""))
# AUC seem to be the best yet AUC = 0.907
# To balance off performance between high True positive and low False positive we can use around 0.25

#Calculate logistic regression complex using cut of 0.425 on test set
cutoff3<-0.425
class.rf <-factor(ifelse(rf.prob>cutoff3,">50K.","<=50K."),levels=c("<=50K.",">50K."))
confusionMatrix(class.rf,adult.test$income)
# Accuracy : 0.8114 Sensitivity : 0.8118 Specificity : 0.8102

###############################################################################################

set.seed(1)
# By default use sqrt(p) number variables when building a random forest of classification trees
# Try Number of Trees to 100
rf.adult1 = randomForest(income~.,data=adult,ntree=100)

importance(rf.adult1)
plot(rf.adult1)
# Run randomForest prediction
rf.pred.prob1 = predict(rf.adult1,newdata=testdata,type = "prob")
rf.prob1 = rf.pred.prob1[,2]

# Generate ROC curve to find a cutoff to use
results.rf1 <-prediction(rf.prob1, adult.test$income,label.ordering=c("<=50K.",">50K."))
roc.rf1 = performance(results.rf1, measure = "tpr", x.measure = "fpr")
plot(roc.rf1,colorize = TRUE)
abline(a=0, b= 1)
auc.rf1 <- performance(results.rf1, measure = "auc")
auc.rf1 <- auc.rf1@y.values
text(x = .40, y = .6,paste("AUC = ", round(auc.rf1[[1]],3), sep = ""))
# AUC seem to be the best yet AUC = 0.902
# To balance off performance between high True positive and low False positive we can use around 0.25

#Calculate logistic regression complex using cut of 0.43 on test set
cutoff4<-0.41
class.rf1 <-factor(ifelse(rf.prob1>cutoff4,">50K.","<=50K."),levels=c("<=50K.",">50K."))
confusionMatrix(class.rf1,adult.test$income)
# Accuracy : 0.8089 Sensitivity : 0.8114 Specificity : 0.8012

#### Overlay multiple ROC curve together
```{r}
plot(roc.sm1)
plot(roc.complex2,col="orange", add = TRUE)
plot(roc.lda,col="blue",add=TRUE)
plot(roc.rf,col="green",add=TRUE)
legend("bottomright",legend=c("Simple LR","Complex LG","LDA","Random Forest"),col=c("black","orange","blue","green"),lty=1,lwd=1)
abline(a=0, b= 1)
